---
title: "Strain Vs. Cases (COVID-19)"
author: "Julie Gilbert"
date: "7/14/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```


Cases data pulled from the State of Michigan's publicly available data. 

Cases include confirmed + probable cases, exclude correctional facilities. Cases are grouped into weeks by onset date. 

Public Health Region 2S = Monroe, Washtenaw, Wayne, Detroit City

Michigan = All counties, excluding correctional facilities & those denoted "out of state", includes unknown county cases.


```{r}
library(tidyverse)
library(lubridate)
library(rvest)
library(readr)
library(openxlsx)
library(gt)

source("save_paths.R")

```

```{r}
#### This section pulls the most recent case and death data from the state of michigan

source("pull_cases_from_michigan_website.R")

```


File name that case data is pulled from: `r paste0(basename(doc_url), ".xlsx")`

```{r}

### read in genomic sequencing data results
data_file <- data_file_source


# filter out anything that suggests a negative control
data_file <- filter(data_file, !grepl("Negat", SampleSourceLocation) & !grepl("HeLa", sample_id) & !grepl("NC_", sample_id))

# filter out non-michigan based project sources
data_file <- filter(data_file, !received_source %in% c("CDCIVY", "CDCIVY4", "RVTN", "BSL3"))

#who_cw <- read.csv("/who_variant_crosswalk.csv")

### add overarching strain name based on pangolin lineage
data_file <- data_file %>% mutate(pangolin_lineage = trimws(pangolin_lineage))
data_file <- data_file %>% mutate(strain_name = case_when(pangolin_lineage == "B.1.1.529" ~ "Omicron",
                                                          grepl("BA.1", pangolin_lineage) ~ "Omicron BA.1 Type", 
                                                          grepl("BA.2", pangolin_lineage) ~ "Omicron BA.2 Type", 
                                                          grepl("BA.3", pangolin_lineage) ~ "Omicron BA.3 Type", 
                                                          grepl("BA.4", pangolin_lineage) ~ "Omicron BA.4 Type", 
                                                          grepl("BA.5", pangolin_lineage) ~ "Omicron BA.5 Type", 
                                                          grepl("AY.", pangolin_lineage) ~ "Delta", 
                                                          pangolin_lineage == "B.1.617.2" ~ "Delta", 
                                                          pangolin_lineage == "B.1.1.7" ~ "Alpha", 
                                                          grepl("Q.", pangolin_lineage) ~ "Alpha",
                                                          pangolin_lineage == "B.1.351" ~ "Beta", 
                                                          pangolin_lineage == "P.1" ~ "Gamma", 
                                                          pangolin_lineage == "B.1.427" ~ "Epsilon",
                                                          pangolin_lineage == "B.1.429" ~ "Epsilon",
                                                          pangolin_lineage == "B.1.525" ~ "Eta",
                                                          pangolin_lineage == "B.1.526" ~ "Iota",
                                                          pangolin_lineage == "B.1.617.1" ~ "Kappa",
                                                          pangolin_lineage == "B.1.621" ~ "Mu",
                                                          pangolin_lineage == "B.1.621.1" ~ "Mu",
                                                          pangolin_lineage == "P.2" ~ "Zeta",
                                                          grepl("B.", pangolin_lineage) ~ "Original B", 
                                                          pangolin_lineage == "B" ~ "Original B",
                                                          pangolin_lineage == "D.2" ~ "Original B",
                                                          pangolin_lineage == "R.1" ~ "Original B",
                                                          pangolin_lineage == "C.37" ~ "Original B",
                                                          grepl("A.", pangolin_lineage) ~ "Original A",
                                                          pangolin_lineage == "A" ~ "Original A",
                                                          pangolin_lineage == "" ~ "No Call",
                                                          pangolin_lineage == "None" ~ "No Call",
                                                          pangolin_lineage == "Unassigned" ~ "No Call",
                                                          T ~ "Unknown"))


#data_file <- merge(data_file, who_cw, by.x = c("pangolin_lineage"), by.y = c("pango_lineage"), all.x = TRUE)

case_data <- read.csv(new_file_name_and_loc)

```

Negative control rows and CDC IVY samples removed from full sample list. 


```{r}

### sorting out case data

# washtenaw county
washtenaw_cases <- filter(case_data, COUNTY == "Washtenaw")

washtenaw_cases$week <- epiweek(washtenaw_cases$Date)
washtenaw_cases$week <- ifelse(nchar(as.character(washtenaw_cases$week)) == 1, paste0("0", as.character(washtenaw_cases$week)), as.character(washtenaw_cases$week)) 

washtenaw_cases$year <- year(washtenaw_cases$Date)

#washtenaw_cases$chart_date <- paste0(washtenaw_cases$year, "-", washtenaw_cases$week)

washtenaw_cases <- washtenaw_cases %>% group_by(year, week) %>% summarize(case_cases = sum(Cases, na.rm = TRUE))

#PH region 2s = monroe, washtenaw, wayne (+ detroit city)
ph_region_2s_cases <- filter(case_data, COUNTY %in% c("Washtenaw", "Monroe", "Wayne", "Detroit City"))

ph_region_2s_cases$week <- epiweek(ph_region_2s_cases$Date)
ph_region_2s_cases$week <- ifelse(nchar(as.character(ph_region_2s_cases$week)) == 1, paste0("0", as.character(ph_region_2s_cases$week)), as.character(ph_region_2s_cases$week)) 

ph_region_2s_cases$year <- year(ph_region_2s_cases$Date)

#ph_region_2s_cases$chart_date <- paste0(ph_region_2s_cases$year, "-", ph_region_2s_cases$week)

ph_region_2s_cases <- ph_region_2s_cases %>% group_by(year, week) %>% summarize(case_cases = sum(Cases, na.rm = TRUE))

colnames(ph_region_2s_cases) <- c("year", "week", "ph_2s_cases")

# full state = no MDOC, no out of state
michigan_cases <- filter(case_data, COUNTY != "MDOC" & COUNTY != "Out-of-State")

michigan_cases$week <- epiweek(as_date(michigan_cases$Date))
michigan_cases$week <- ifelse(nchar(as.character(michigan_cases$week)) == 1, paste0("0", as.character(michigan_cases$week)), as.character(michigan_cases$week)) 

michigan_cases$year <- year(as_date(michigan_cases$Date))

#michigan_cases$chart_date <- paste0(michigan_cases$year, "-", michigan_cases$week)

michigan_cases <- michigan_cases %>% group_by(year, week) %>% summarize(case_cases = sum(Cases, na.rm = TRUE))


```

Populations - 2019 Numbers, from the State of Michigan: 

Monroe: 150500
Wayne: 1079291
Washtenaw: 367601
Detroit City: 670052

PH Region 2S = 2267443

```{r}
#data_file$WHO_LABEL <- ifelse(is.na(data_file$WHO_LABEL), "Not VOI/VOC", data_file$WHO_LABEL)

data_file$counter <- 1

data_file <- data_file %>% mutate(coll_date = case_when(grepl("/", coll_date) ~ as.POSIXct(coll_date, format = "%m/%d/%Y"), 
                                                        grepl("-", coll_date) ~ as.POSIXct(coll_date, format = "%Y-%m-%d"), 
                                                        T ~ NA_real_))

data_file$collection_week <- epiweek(data_file$coll_date)
data_file$collection_week <- ifelse(nchar(as.character(data_file$collection_week)) == 1, paste0("0", as.character(data_file$collection_week)), as.character(data_file$collection_week)) 

data_file$collection_year <- year(data_file$coll_date)

# week is counting last week of 2020 as week "53"
### adjust years and weeks
data_file <- data_file %>% mutate(collection_week = case_when(collection_week == "53" 
                                                     & collection_year == "2021" ~ "01", 
                                                     collection_week == "52" 
                                                     & collection_year == "2022" ~ "01",
                         T ~ collection_week))

# filter out everything that is from 2020
data_file <- filter(data_file, collection_year != 2020)

#data_file$chart_collection_date <- paste0(data_file$collection_year, "-", data_file$collection_week)


###########################################

data_file90 <- filter(data_file, nextclade_completeness >= 90)

strain_totals90 <- data_file90 %>% group_by(collection_year, collection_week, strain_name) %>% summarize(strain_count = sum(counter, na.rm = TRUE))

totals90 <- data_file90 %>% group_by(collection_year, collection_week) %>% summarize(total = sum(counter, na.rm = TRUE))


strain_totals90 <- merge(strain_totals90, totals90, by = c("collection_year", "collection_week"), all = TRUE)




###########################################

strain_totals <- data_file %>% group_by(collection_year, collection_week, strain_name) %>% summarize(strain_count = sum(counter, na.rm = TRUE))

totals <- data_file %>% group_by(collection_year, collection_week) %>% summarize(total = sum(counter, na.rm = TRUE))


strain_totals <- merge(strain_totals, totals, by = c("collection_year", "collection_week"), all = TRUE)

```


#### Charts - All Sequences 

```{r, fig.height = 8, fig.width = 10}

strain_totals <- merge(strain_totals, ph_region_2s_cases, by.x = c("collection_year", "collection_week"), by.y = c("year", "week"), all.x = TRUE)

# turn this to cases per 100,000 population
strain_totals$ph_2s_cases_per100 <- round((strain_totals$ph_2s_cases / 2267443) * 100000, 1)

ggplot(strain_totals) + 
  geom_bar(stat = "identity", aes(x = collection_week, y = strain_count, fill = strain_name)) + 
  geom_line(data = strain_totals, aes(x = collection_week, y = ph_2s_cases_per100), color = "black", group = 1, alpha = 0.6, size = 1) +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  scale_fill_manual(values = c("#F2929A", "#392061", "#92817A", "#1C3041", "#0BC18A", 
                               "#006989", "#ACB0BD", "#FF6663", "#85D4FF",  
                               "#87C38F", "#48BEFF", "#AA6373", "#F0F2A6", 
                               "#DA7422", "#62466B", "#FCEFF9")) + 
  labs(title = "Samples Collected per Week", 
       subtitle = "Colored by Lineage Name", 
       x = "Collection Week", 
       y = "Count", 
       fill = "Lineage Name", 
       caption = "Black line = New Cases per 100,000 population in Public Health Region 2S") + 
  facet_wrap(.~collection_year, ncol = 1)

```


```{r, fig.height = 8, fig.width = 10}
strain_totals$proportion <- round((strain_totals$strain_count / strain_totals$total)*100, 1)

ggplot(strain_totals) + 
  geom_bar(stat = "identity", aes(x = collection_week, y = proportion, fill = strain_name)) + 
  geom_line(data = strain_totals, aes(x = collection_week, y = ph_2s_cases_per100/10), color = "gray40", group = 1, alpha = 0.6, size = 2) +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  scale_fill_manual(values = c("#F2929A", "#392061", "#92817A", "#1C3041", "#0BC18A", 
                               "#006989", "#ACB0BD", "#FF6663", "#85D4FF",  
                               "#87C38F", "#48BEFF", "#AA6373", "#F0F2A6", 
                               "#DA7422", "#62466B", "#FCEFF9")) +  
  labs(title = "Samples Collected per Week - Proportions of Each Strain", 
       subtitle = "Colored by Lineage Name", 
       x = "Collection Week", 
       y = "Proportion (%)", 
       fill = "Lineage Name", 
       caption = "Gray line = New Cases per 100,000 population in Public Health Region 2S divided by 10") + 
  facet_wrap(.~collection_year, ncol = 1)

```

#### Table of Most Recent 50 Samples

Ordered samples by collection date, most recent 50 samples selected.
Only samples with > 90% nextclade completeness shown

```{r}

data_file_order <- filter(data_file, !is.na(coll_date) & nextclade_completeness > 90) %>% arrange(coll_date)
data_file_50 <- tail(data_file_order, 50)

df50 <- as.data.frame(table(data_file_50$nextclade_clade, data_file_50$pangolin_lineage, useNA = "always"))

colnames(df50) <- c("NextClade", "Pangolin", "Count")
df50 <- filter(df50, Count != 0)
df50 <- df50 %>% arrange(desc(Count))

df50 %>% 
  gt() %>% 
  tab_header(
    title = "Most Recent 50 Samples",
    subtitle = paste0("Collected ", min(data_file_50$coll_date), " to ", max(data_file_50$coll_date))
  ) 
```


---

#### Charts - Sequences with >= 90% Completeness


```{r, fig.height = 8, fig.width = 10}

strain_totals90 <- merge(strain_totals90, ph_region_2s_cases, by.x = c("chart_collection_date"), by.y = c("chart_date"), all.x = TRUE)

# turn this to cases per 100,000 population
strain_totals90$ph_2s_cases_per100 <- round((strain_totals90$ph_2s_cases / 2267443) * 100000, 1)

ggplot(strain_totals90) + 
  geom_bar(stat = "identity", aes(x = as.character(chart_collection_date), y = strain_count, fill = WHO_LABEL)) + 
  geom_line(data = strain_totals90, aes(x = as.character(chart_collection_date), y = ph_2s_cases_per100), color = "black", group = 1, alpha = 0.6, size = 1) +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  scale_fill_manual(values = c("#F2929A", "#92817A", "#1C3041", "#0BC18A", 
                               "#006989", "#FF6663", "#85D4FF", "#ACB0BD", "#48BEFF")) + 
  labs(title = "Samples Collected per Week", 
       subtitle = "Colored by WHO Label", 
       x = "Collection Week", 
       y = "Count", 
       fill = "WHO Label", 
       caption = "Black line = New Cases per 100,000 population in Public Health Region 2S") 


```

```{r, fig.height = 8, fig.width = 10}

strain_totals90$proportion <- round((strain_totals90$strain_count / strain_totals90$total)*100, 1)

ggplot(strain_totals90) + 
  geom_bar(stat = "identity", aes(x = as.character(chart_collection_date), y = proportion, fill = WHO_LABEL)) + 
  geom_line(data = strain_totals90, aes(x = as.character(chart_collection_date), y = ph_2s_cases_per100/4), color = "black", group = 1, alpha = 0.6, size = 1) +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  scale_fill_manual(values = c("#F2929A", "#92817A", "#1C3041", "#0BC18A", 
                               "#006989", "#FF6663", "#85D4FF", "#ACB0BD", "#48BEFF")) + 
  labs(title = "Samples Collected per Week - Proportions of Each Strain", 
       subtitle = "Colored by WHO Label", 
       x = "Collection Week", 
       y = "Proportion (%)", 
       fill = "WHO Label", 
       caption = "Black line = New Cases per 100,000 population in Public Health Region 2S divided by 4") 

```


```{r}
data_file_order <- filter(data_file90, !is.na(coll_date)) %>% arrange(coll_date)
data_file_90_50 <- tail(data_file_order, 50)

df90_50 <- as.data.frame(table(data_file_90_50$nextclade_clade.x, data_file_90_50$pangolin_lineage, useNA = "always"))

colnames(df90_50) <- c("NextClade", "Pangolin", "Count")
df90_50 <- filter(df90_50, Count != 0)
df90_50 <- df90_50 %>% arrange(desc(Count))

df90_50 %>% 
  gt() %>% 
  tab_header(
    title = "Most Recent 50 Samples",
    subtitle = paste0("Collected ", min(data_file_90_50$coll_date), " to ", max(data_file_90_50$coll_date))
  ) 

```

